Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-05-03 Paper/Manuscript.tex
Corrections up to page 10 and one new reference.

\author{Sarp Akcay}
\affiliation{School of Mathematics \& Statistics, University College Dublin, Belfield, Dublin 4, Ireland.}
\affiliation{Friedrich-Schiller-Universit{\"a}t Jena}
%\date{\today}

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars sufficiently
early to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very beginning.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the inteferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time, thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times of up to five hours.  
Taking one hour as a benchmark warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS event rates within this horizon distance, we show that this translates to forecasts of $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
This article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
Though many of our results may be known to the experts, they might nonetheless find this article motivating and exciting.
\end{abstract}
\maketitle

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $ 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched the largest-ever campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain threshold signal-to-noise ratio.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

We compute the advance warning times via the following procedure: 
(${i}$) We start with a GW source presumed to be an inspiralling BNS at a certain luminosity distance. %$D$.
(${ii}$) We determine the frequency at which the source enters a given interferometer's bandwidth. 
%which reduce an individual interferometer's sensitivity to a particular source. 
(${iii}$) We determine the detection time by computing the frequency at which the network's accumulated signal-to-noise ratio (SNR) for the BNS inspiral equals 15.
($iv$) We define the advance warning time, $T_\text{AW}$, to be the remaining inspiral time to the merger from the instant of detection.
We repeat this procedure for BNSs at luminosity distances varying from 50\,Mpc to 1\,Gpc with the smaller values intended for the ALV network and 
the larger ones for Einstein Telescope.
In order to more faithfully represent an interferometer network such as ALV --- consisting of three separate interferometers which have different
orientations and positions --- we employ root-mean-square averages over sky-position and polarization angles,
and assign the interferometers angle-averaged sensitivities.
%KAGRA 152Mpc range (K. Somiya, ET meeting at Birmingham, 2017)

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment formulated in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Baiotti:2016qnr,Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simple formulation, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

%Let us emphasize that this article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
%Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
%Be that as it may, even the experts might find our results motivating and exciting.
%
This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020} and \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $t$ denotes observer/detector time and $f$ denotes the GW frequency of the dominant quadrupole mode, which is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments. Details of how to obtain this expression can be found in Ref.~\cite{Maggiore}.

Here, we are specifically interested in the motion of two point masses in a circular orbit around a common center of mass.
Labelling the masses by $m_1 \le m_2$, we have, for a binary in a circular orbit with Keplerian angular frequency $\Omega$ and separation $r$,
%
\be
\dot{E} = \f{32}{5}\f{G\mu^2}{c^5}\, r^4 \Omega^6\label{eq:Edot},
\ee
%
where $\mu= m_1 m_2/M$ is the reduced mass and $M=m_1+m_2$ is the total mass. 
We justify our simplification of the motion to circular orbits in Sec.~\ref{sec:idealizations}.
%Our assumption of the circularity of the motion is well justified in

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}= \f{32}{5}\f{c^5}{G}x^5, \label{eq:Edot2}
\ee
%
where $x = ({G M_c\, \Omega}/{c^3})^{2/3} $ is the dimensionless inverse separation parameter ubiquitous in post- Newtonian theory and $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from Eq.~(\ref{eq:Edot2}) that $\dot{E}\propto x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,.  \label{eq:binding_energy1}
\ee
%
$\dot{E}_b \propto \dot{r}/r^2$, so $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases
making the orbit more bound in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
Using $\tau$ we can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$. This is the canonical value for the 
mass of a neutron star due to Chandrasekhar limit \cite{Chandra}.

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular (ISCO) orbit in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim \ord(10^3)\,$Hz for a system with $M =2.8 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations are $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

@@

use design sensitivities of A-LIGO and ET's B and C configurations for $\sqrt{S_n(f)}$.

>>>>

\author{Sarp Akcay}
\affiliation{School of Mathematics \& Statistics, University College Dublin, Belfield, Dublin 4, Ireland}
\affiliation{Theoretisch-Physikalisches Institut, Friedrich-Schiller-Universit{\"a}t Jena, 07743, Jena, Germany}
%\date{\today}

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars sufficiently
early to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very beginning.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the inteferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time, thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times of more than five hours for $D \le 100\,$Mpc.
Taking one hour as a benchmark advance warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS merger event rates within this volume, we show that Einstein C will forecast $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
This article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
Though many of our results may be known to the experts, they might nonetheless find this article motivating and exciting.
\end{abstract}
\maketitle

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $ 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched a massive campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain detection criterion which we describe below.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

We compute the advance warning times via the following procedure: 
(${i}$) We start with a GW source presumed to be an inspiralling BNS at a certain luminosity distance. %$D$.
(${ii}$) We determine the frequency at which the source enters a given interferometer's bandwidth. 
%which reduce an individual interferometer's sensitivity to a particular source. 
(${iii}$) We determine the detection time by computing the frequency at which the network's accumulated signal-to-noise ratio for the BNS inspiral equals 15.
($iv$) We define the advance warning time, $T_\text{AW}$, to be the time interval between the instant of detection and the merger.
We repeat this procedure for BNSs at luminosity distances varying from 50\,Mpc to 1\,Gpc with the smaller values intended for the ALV network and 
the larger ones for Einstein Telescope.
In order to more faithfully represent an interferometer network such as ALV --- consisting of three separate interferometers which have different
orientations and positions --- we employ root-mean-square averages over sky-position and polarization angles,
and assign the interferometers angle-averaged sensitivities.
%KAGRA 152Mpc range (K. Somiya, ET meeting at Birmingham, 2017)

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Baiotti:2016qnr,Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simple formulation, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

%Let us emphasize that this article is intended as a pedagogical introduction to gravitational-wave astronomy written at a level accessible to
%Ph.D. students, advanced undergraduates, and colleagues in astronomy and/or astrophysics who wish to learn more about the underlying physics.
%Be that as it may, even the experts might find our results motivating and exciting.
%
This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020} and \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of binary black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $t$ denotes observer/detector time and $f$ denotes the GW frequency of the dominant quadrupole mode in observer/detector frame. 
%$f$ is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments. Details of how to obtain this expression can be found in Ref.~\cite{Maggiore}.
There is no monopole radiation due to conservation of mass and no dipole radiation due to conservation of linear and angular momenta
(cf. Ref.~\cite{Mono}).

Here, we are specifically interested in the motion of two point masses in a circular orbit around a common center of mass.
For a binary in a circular orbit with Keplerian angular frequency $\Omega$, separation $r$, and masses $m_1 \le m_2$, we have \cite{Maggiore}
%
\be
\dot{E} = \f{32}{5}\f{G\mu^2}{c^5}\, r^4 \Omega^6\label{eq:Edot},
\ee
%
where $\mu= m_1 m_2/M$ is the reduced mass and $M=m_1+m_2$ is the total mass. 
We justify our simplification of the motion to circular orbits in Sec.~\ref{sec:idealizations}.
%Our assumption of the circularity of the motion is well justified in

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}= \f{32}{5}\f{c^5}{G}x^5, \label{eq:Edot2}
\ee
%
where $x = ({G M_c\, \Omega}/{c^3})^{2/3} $ is the dimensionless inverse separation parameter ubiquitous in post- Newtonian theory and $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from Eq.~(\ref{eq:Edot2}) that 
at the leading order, power emitted in GWs scales as $x^5$.
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,.  \label{eq:binding_energy1}
\ee
%
$\dot{E}_b \propto \dot{r}/r^2$, so $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases
making the orbit more bound in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
Using $\tau$ we can now define the inspiral time, i.e, time left to coalesence at a given GW frequency $f$
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$. This is the canonical value for the 
mass of a neutron star due to the Chandrasekhar limit \cite{Chandra}.

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular orbit (ISCO) in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim \ord(10^3)\,$Hz for a system with $M =2.8 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations are $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

@@

use design sensitivities of A-LIGO and ET for $\sqrt{S_n(f)}$.
